# Introduction to R and Bioconductor for Cell Biology 8401

## R basics

### Basic operations

R as a glorified calculator. Try typing this in directly into the console. Then start typing this into the editor, and save your script. Use the run button, or press `CMD`+`Enter` (`Ctrl`+`Enter` on Windows).

```{r, results='hide'}
2+2
5*4
2^3
```

R Knows order of operations.

```{r, results='hide'}
2+3*4
```

### Functions

R has built-in functions.

```{r, results='hide'}
# Notice that this is a comment. 
# Anything behind a # is "commented out" and is not run.
sqrt(144)
log(1000)
```

Get help by typing a question mark in front of the function's name:

```
?log
```

Note syntax highlighting when typing this into the editor. Also note how we pass *arguments* to functions. Finally, see how you can *next* one function inside of another (here taking the square root of the log-base-10 of 1000).

```{r, results='hide'}
log(1000)
log(1000, base=10)
sqrt(log(1000, base=10))
```

### Vectors

Let's create some numeric vectors. Vectors (aka "arrays" in Perl, "lists" in Python) are single *objects* containing an ordered collection of *elements*. A simple vector is a numeric vector, a single *object* containing several numbers. Here let's display a few vectors. We can also do vector arithmetic. When printing vectors to the screen that have lots of elements, notice that the bracketed number in the gutter of the output is just a counter indexing the number of elements in the vector.

```{r, eval=FALSE}
# Some simple numeric vectors:
1:5
6:10
1:5 + 6:10  
1:100 

# Get some help with the seq() function, then create a vector from 2 to 200 by 2s.
# Notice how the seq() function works -- the `to` argument will never be exceeded.
?seq
seq(from=2, to=200, by=4)
```

### Variables

You can store values in a variable. Use the `<-` operator for assignment. `=` also will work, but `<-` is conventional and preferred. Named variables are objects containing whatever is assigned to them. 

```{r, results='hide'}
x <- 5
x

y <- 42
y

y-x
z <- y-x
z

x <- 1:5
y <- 6:10
x
y
x+y
x*y
x^y
```

You can see what objects (variables) are stored by viewing the Environment tab in Rstudio. You can also use the `ls()` function. You can remove objects (variables) with the `rm()` function. You can do this one at a time or remove several objects at once.

```{r, results='hide'}
ls()
rm(x)
ls()
x # oops! you should get an error because x no longer exists!
rm(y,z)
```

### Classes: everything is an object

Use the `class()` function to see what *kind* of object a something is. You can run `class()` on constants, built-in objects, or objects you create. Let's create a character object and then get it's class. 

```{r, results='hide'}
class(42)
class(log)
name <- "Stephen"
name
class(name)
```

Certain *functions* operate only on certain *classes* of object. Here, `name` is a `character` class, assigned to `"Stephen"`. The built-in `toupper()` function will operate on character objects, but not others. 

```{r}
toupper(name) # name is an object of class character. methods or functions are associated with certain classes.
toupper(log) # can't run a function that expects character on an object of class function
```

We can combine values into a vector with the built-in `c()` function.

```{r, results='hide'}
# Get some help with ?c
x <- c(1,3,5)
x
class(x)
length(x)
```

Let's create and manipulate a character vector:

```{r, results='hide'}
y <- c("My", "name", "is", "Stephen")
y
class(y)
length(y)
y <- c(y, "Turner")
y
length(y)
```

Try running the built-in `sum()` function on a numeric vector:

```{r, results='hide'}
sum(x)
```

Combining characters with numerics results in coercing everything to be a character class. 

```{r, results='hide'}
y
z <- c(x,y)
z 
class(z)
```

Certain functions only operate on certain classes. You can't compute the `sum()` of a character vector!

```{r}
z
sum(z)
```

### Subsetting

Lorem ipsum.


```{r, eval=FALSE, echo=FALSE}
# subsetting
x <- 101:150
x
x[1]
x[50]
x[20:30]
x[40:60] #NA is missing value!


# data frames
# load motor trend car road tests from 1974
data(mtcars)
mtcars
head(mtcars)
length(mtcars) # the number of features
dim(mtcars)
dim(mtcars)[1] # number of cars (samples)
str(mtcars)

#access elements
mtcars$mpg
mtcars$cyl
# subset
head(mtcars)
mtcars[1:4, 1:2]
# see only 6 cylinder cars, etc
subset(mtcars, cyl==6)
subset(mtcars, cyl>6)
subset(mtcars, cyl==6, select=c(mpg, disp))

# with
mtcars$mpg / mtcars$cyl * mtcars$disp
with(mtcars, mpg/cyl*disp)

# plot
plot(mtcars$mpg)
with(mtcars, plot(mpg))
hist(mtcars$mpg)
hist(mtcars$mpg, breaks=10)
hist(mtcars$mpg, breaks=10, col="black")
hist(mtcars$mpg, breaks=10, col="gray50")
plot(mtcars$disp, mtcars$mpg)
with(mtcars, plot(disp, mpg))
with(mtcars, plot(disp, mpg, col="green"))
with(mtcars, plot(disp, mpg, pch=16))
with(mtcars, plot(disp, mpg, pch=16,  col="red"))
with(mtcars, plot(disp, mpg, pch=16,  col="red", main="MPG vs Displacement"))
with(mtcars, plot(disp, mpg, pch=16,  col="red", main="MPG vs Displacement", 
                  ylab="Miles per Gallon", xlab="Displacement (cu. in.)"))

# exercise: plot horsepower vs displacement only for 8 cylinder vehicles. 
# give it a title, label the x and y axes
with(subset(mtcars, cyl==8), plot(disp, hp, pch=16, xlab="Displacement", ylab="Horsepower", main="HP vs Disp"))
with(mtcars, plot(disp, hp, pch=16, xlab="Displacement", ylab="Horsepower", main="HP vs Disp"))

# saving
mtcars_8cyl <- subset(mtcars, cyl==8)
mtcars_8cyl
?write.table
?write.csv
getwd()
# go to session->set working directory, change to desktop/r
getwd()
setwd("/Users/sdt5z/Desktop/R")
write.csv(mtcars_8cyl, file="cars8.csv")

## remove the dataset, load again using rstudio tools menu, or manual command
rm(mtcars_8cyl)
mtcars_8cyl
?read.csv
cars8 <- read.table(file="cars8.csv", header=TRUE, sep=",", row.names=1)


# BioC --------------------------------------------------------------------

## http://www.bioconductor.org/packages/2.11/bioc/vignettes/SPIA/inst/doc/SPIA.pdf
## Download data from http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE4107
## http://people.virginia.edu/~sdt5z/GSE4107_RAW.zip
## Extract tar file, gunzip all the .CEL.gz files

# Do this in folder on desktop
setwd("~/R")

biocLite(c("affy", "AnnotationDbi", "hgu133plus2cdf", "hgu133plus2.db", "genefilter", "DBI", "annotate", "arrayQualityMetrics", "limma", "GOstats", "Category", "GO.db", "KEGG.db")

?list.celfiles
library(affy)
?list.celfiles
list.celfiles("GSE4107_RAW", full.names=T)
myfiles <- list.celfiles("GSE4107_RAW", full.names=T)

?ReadAffy
myaffybatch <- ReadAffy(filenames=myfiles)
myaffybatch
class(myaffybatch)
?"AffyBatch-class"

?rma
eset <- rma(myaffybatch)
class(eset)
str(eset)
?"ExpressionSet-class"
head(exprs(eset))
dim(exprs(eset))

## Annotate the samples
pData(eset)
class(pData(eset))
pData(eset)$condition
rep("cancer", 12)
rep("healthy", 10)
c(rep("cancer", 12), rep("healthy", 10))
factor(c(rep("cancer", 12), rep("healthy", 10)))
pData(eset)$condition <- factor(c(rep("cancer", 12), rep("healthy", 10)))
pData(eset)


# filter the dataset
library(genefilter)
eset
?featureFilter
eset <- featureFilter(eset, require.entrez=T, remove.dupEntrez=T, feature.exclude="^AFFX")
eset

## Annotate the features
fData(eset)
annotation(eset)
library(annotate)
library(hgu133plus2.db)
ls("package:hgu133plus2.db")
ID     <- featureNames(eset)
class(ID)
head(ID)
head(exprs(eset))
?lookUp
#trust me on this one
Symbol <- as.character(lookUp(ID, "hgu133plus2.db", "SYMBOL"))
Name   <- as.character(lookUp(ID, "hgu133plus2.db", "GENENAME"))
Entrez <- as.character(lookUp(ID, "hgu133plus2.db", "ENTREZID"))
tmp <- data.frame(ID=ID, Entrez=Entrez, Symbol=Symbol, Name=Name, stringsAsFactors=F)
head(tmp)
fData(eset) <- tmp
head(fData(eset))
rm(ID, Symbol, Name, Entrez, tmp)

# Quality Assessment ------------------------------------------------------

# Load the arrayQualityMetrics package
library(arrayQualityMetrics)

# Run arrayQualityMetrics on your eset, giving it the name of an ouput directory.
# intgroup is the name of the sample covariate(s) used to draw a colour side bar next to the heatmap. character matching column in pData(eset)
# also give it a title
?arrayQualityMetrics
arrayQualityMetrics(eset, outdir="aqm", intgroup="condition", reporttitle="Quality Assessment Report")
# Now go open the current working directory, and open aqm/index.html in your browser


# Data Analysis -----------------------------------------------------------

## The limma package will be used for analysis
library(limma)

## Make the design and contrasts matrix
pData(eset)
design <- model.matrix(~0+condition, data=pData(eset))
design
class(design)
colnames(design)
colnames(design) <- c("cancer", "healthy")
design
colnames(design) <- sub("condition", "", colnames(design))
design

## make contrast matrix
contrast.matrix <- makeContrasts(cancer_v_healthy=cancer-healthy, levels=design)
contrast.matrix

## Fit model
fit <- lmFit(eset,design)
fit <- contrasts.fit(fit, contrast.matrix)
fit <- eBayes(fit)
class(fit)

## Get statistics from that model fit
topTable(fit)
?topTable
nrow(fit)
tt <- topTable(fit, number=nrow(fit))

head(tt)
class(tt)
dim(tt)

## First, write the entire results table to a csv file and open in Excel
## Assignment: how many genes are significant at an adjusted p-value of 0.005?
## Write this table to a CSV file and open it in excel
write.csv(tt, file="diffexpr-all.csv")
sig <- subset(tt, adj.P.Val<0.005)
dim(sig)
write.csv(sig, file="diffexpr-significant.csv")

## volcano plot
with(tt, plot(logFC, -log10(adj.P.Val), pch=16, col="black"))

## heatmap
heatmap(exprs(eset[featureNames(eset) %in% sig$ID, ]))

library(GOstats)
library(Category)
?hyperGTest
?"GOHyperGParams-class"
myuniverse <- tt$Entrez
mysiggenes <- sig$Entrez
myanno <- annotation(eset)

?"KEGGHyperGParams-class"
library(KEGG.db)
KEGGparams <- new("KEGGHyperGParams", geneIds=mysiggenes, universeGeneIds=myuniverse, annotation=myanno, testDirection="over", pvalueCutoff=.05)
KEGGres <- hyperGTest(KEGGparams)
KEGGres
class(KEGGres)
?"KEGGHyperGResult-class"
?"HyperGResult-accessors"
summary(KEGGres)
htmlReport(KEGGres, file="KEGG report.html")

?"GOHyperGParams-class"
library(GO.db)
GOparams <- new("GOHyperGParams", geneIds=mysiggenes, universeGeneIds=myuniverse, annotation=myanno, conditional=TRUE, ontology="MF")
GOres <- hyperGTest(GOparams)
htmlReport(GOres, file="GO MF terms.html")
```